### 2025-08-26 sections/2-methods.tex
partial methods edit

In this manuscript, we are primarily concerned with $k$-NN search in a finite-dimensional space.
Given a dataset $\textbf{X} = \{x_1 \dots x_n\}$ of cardinality $|\textbf{X}| = n$, we define a \textit{point} or \textit{datum} $x_i \in \textbf{X}$ as a singular observation. Examples include the neural-network embedding of an image, the genome of an organism, and a measurement of a radio frequency signal.

We define a \textit{distance function} $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \ \cup \ \{0\}$ which, given two points, deterministically returns a finite and non-negative real number.
A distance value of zero defines an identity among points (i.e.,\,$f(x, y) = 0 \Leftrightarrow x = y$) and larger values indicate greater dissimilarity among points.
We also require that the distance function be symmetric (i.e.,\,$f(x, y) = f(y, x) \ \forall \ x, y \in \textbf{X}$).
In addition to these constraints, if the distance function obeys the triangle inequality (i.e.,\,$f(x, y) \leq f(x, z) + f(z, y) \ \forall \ x, y, z \in \textbf{X}$), then it is also a \textit{distance metric}.
Similar to~\cite{yu2015entropy}, when used with distance metrics, all search algorithms in CAKES are exact.
For example, Euclidean, Levenshtein~\cite{levenshtein1966binary} and Dynamic Time Warping (DTW)~\cite{muller2007dynamic} distances are all distance metrics, while cosine distance is not a metric because it violates the triangle inequality (e.g.,\,consider the points $x = (1, 0)$, $y = (0, 1)$ and $z = (1, 1)$ on the Cartesian plane).

The choice of an appropriate distance function varies by dataset and domain.
For example, with neural-network embeddings, one could use Euclidean or Cosine distance.
With genomic or proteomic sequence data, Levenshtein and Hamming distances are useful.
With time-series data, one could use Dynamic Time Warping (DTW) or Wasserstein distance.

CAKES derives its advantages from the manifold hypothesis~\cite{fefferman2016testing}, the notion that high-dimensional data collected from constrained generating phenomena typically only occupy a low-dimensional manifold within their embedding space.
We say that such data are \textit{manifold-constrained} and have low \textit{local fractal dimension} (LFD).
In other words, we assume that the dataset is embedded in a $D$-dimensional space, but that the data only occupy a $d$-dimensional manifold, where $d \ll D$.
While we sometimes use Euclidean notions to describe the geometric and topological properties of the clusters and manifold, CLAM and CAKES do not rely on such notions;
they serve merely as convenient and intuitive vocabulary to discuss the underlying mathematics.
CAKES exploits the low LFD of such datasets to accelerate search.
We define LFD at some length scale around a point in the dataset as:
\begin{equation}
    \text{LFD}(q, r_1, r_2) = \frac{\text{log} \left( \frac{|B(q, r_1)|}{|B(q, r_2)|} \right) }{\text{log} \left( \frac{r_1}{r_2} \right) }
    \label{eq:methods:lfd-original}
\end{equation}
where $B(q, r)$ is the set of points contained in the metric ball of radius $r$ centered at a point $q$ in the dataset $\textbf{X}$.

We use a simplified version of Equation~\ref{eq:methods:lfd-original} by using a length scale where $r_1 = 2 \cdot r_2$.
\begin{equation}
    \text{LFD}(q, r) = \text{log}_2 \left( \frac{|B(q, r)|}{|B(q, \frac{r}{2})|} \right)
    \label{eq:methods:lfd-half}
\end{equation}

>>>>

In this manuscript, we study $k$-NN search in finite-dimensional spaces. Given a dataset $\textbf{X} = \{x_1, \dots, x_n\}$, we refer to each $x_i \in \textbf{X}$ as a point or datum. Examples include image embeddings, genomic sequences, and time-series measurements.

We define a \textit{distance function} $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \ \cup \ \{0\}$ which, given two points, deterministically returns a finite and non-negative real number.
We require $f$ to be symmetric and identity-preserving ($f(x, y) = 0 \iff x = y$). 
If $f$ satisfies the triangle inequality (i.e.,\,$f(x, y) \leq f(x, z) + f(z, y) \ \forall \ x, y, z \in \textbf{X}$), then it is also a \textit{distance metric}. Common examples include Euclidean, Levenshtein~\cite{levenshtein1966binary}, and DTW~\cite{muller2007dynamic}; cosine distance, which violates the triangle inequality, is not a metric.
Similar to~\cite{yu2015entropy}, when used with distance metrics, all search algorithms in CAKES are exact.

% cosine distance is not a metric because it violates the triangle inequality (e.g.,\,consider the points $x = (1, 0)$, $y = (0, 1)$ and $z = (1, 1)$ on the Cartesian plane).

The appropriate distance function depends on the data: Euclidean or cosine is often used for neural embeddings, Levenshtein or Hamming for sequences, and DTW or Wasserstein for time-series data.

CAKES derives its advantages from the manifold hypothesis~\cite{fefferman2016testing}, the notion that high-dimensional data collected from constrained generating phenomena typically only occupy a low-dimensional manifold within their embedding space. We define the \emph{local fractal dimension} (LFD) to quantify this property.

For a point $q \in \textbf{X}$ and radii $r_1 > r_2$,

\begin{equation}
    \text{LFD}(q, r_1, r_2) = \frac{\text{log} \left( \frac{|B(q, r_1)|}{|B(q, r_2)|} \right) }{\text{log} \left( \frac{r_1}{r_2} \right) }
    \label{eq:methods:lfd-original}
\end{equation}
where $B(q, r)$ is the set of points in the metric ball of radius $r$ centered at $q$.

We typically use a simplified version of Equation~\ref{eq:methods:lfd-original} with $r_1 = 2 \cdot r_2$:
\begin{equation}
    \text{LFD}(q, r) = \text{log}_2 \left( \frac{|B(q, r)|}{|B(q, \frac{r}{2})|} \right).
    \label{eq:methods:lfd-half}
\end{equation}

% In other words, we assume that the dataset is embedded in a $D$-dimensional space, but that the data only occupy a $d$-dimensional manifold, where $d \ll D$.
% While we sometimes use Euclidean notions to describe the geometric and topological properties of the clusters and manifold, CLAM and CAKES do not rely on such notions;
% they serve merely as convenient and intuitive vocabulary to discuss the underlying mathematics.
% CAKES exploits the low LFD of such datasets to accelerate search.
